{\bf Affine.m} -- {\it Mathematica} package for computations in representation theory of finite-dimensional and affine Lie algebras

### 2012-06-17 doc/paper.tex
Grammar

Consider an action of the element $s_{\alpha}s_{\alpha+\delta}$ of the Weyl group of affine Lie algebra  $\hat\gf$ for $\alpha$ being a simple root of the underlying finite-dimensional Lie algebra $\gf$. Using definition \eqref{eq:8} it is easy to see that $s_{\alpha}s_{\alpha+\delta} \cdot \lambda=\lambda+\frac{2}{(\alpha,\alpha)}\alpha+\left(\frac{(\alpha,\alpha)}{2 (\lambda,\delta)}+(\lambda,\alpha)\right) \delta$. So the Weyl group $W_{\gfh}$ can be presented as a semidirect product of the Weyl group $W_{\gf}$ of $\gf$ and the set of translations corresponding to the roots of $\gf$.

@@

For a subalgebra $\af\subset \gf$ we introduce the subalgebra
$\afb$. Consider a root subspace $\hf_{\perp \af}^{\ast }$
orthogonal to $\hf_{\af}$,
\begin{equation*}
\hf_{\perp \af}^{\ast }:=\left\{ \eta \in \hf^{\ast }|\forall
h\in \hf_{\af};\eta \left( h\right) =0\right\} ,
\end{equation*}
and the roots (correspondingly -- positive roots) of $\gf$ orthogonal
to the roots of $\af$,
\begin{eqnarray}
\Delta _{\afb} &:&=\left\{ \beta \in \Delta _{\gf}|\forall
h\in \hf_{\af};\beta \left( h\right) =0\right\} ,
\label{delta a ort} \\
\Delta _{\afb}^{+} &:&=\left\{ \beta ^{+}\in \Delta _{\gf%
}^{+}|\forall h\in \hf_{\af};\beta ^{+}\left( h\right) =0\right\} .
\notag
\end{eqnarray}
Let $W_{\afb}$ be a subgroup of $W$ generated by
reflections $w_{\beta }$ with the roots $\beta \in \Delta _{\af_{\perp
}}^{+}$. The subsystem $\Delta _{\afb}$ determines a 
subalgebra $\afb$ with the Cartan subalgebra $\hf_{\af%
_{\perp }}$.

@@

Denote by  $k_{\xi }^{\left( \mu \right) }$ signed branching coefficients. If $\xi\in \bar C_{\af}$ is in main Weyl chamber $k_{\xi}^{(\mu)}=b^{(\mu)}_{\xi}$ otherwise $k_{\xi}^{(\mu)}=\epsilon(w) b^{(\mu)}_{w (\xi+\rho_{\af})-\rho_{\af}}$ where $w\in W_{\af}$ is such that $w (\xi+\rho_{\af})-\rho_{\af}\in \bar C_{\af}$.

Now we can use the Weyl character formula to write a
recurrent relation \cite{2010arXiv1007.0318L} for signed branching
coefficients $k_{\xi }^{\left( \mu \right) }$ corresponding to an
injection $\af\hookrightarrow \gf$:
\begin{equation}
\begin{array}{c}
k_{\xi }^{\left( \mu \right) }=-\frac{1}{s\left( \gamma _{0}\right) }\left(
\sum_{u\in U}\epsilon (u)\;\dim \left( L_{\afb}^{\mu _{\af%
_{\perp }}\left( u\right) }\right) \delta _{\xi -\gamma _{0},\pi _{%
\widetilde{\af}}(u(\mu +\rho )-\rho )}+\right.  \\
\left. +\sum_{\gamma \in \Gamma _{\widetilde{\af}\rightarrow \gf%
}}s\left( \gamma +\gamma _{0}\right) k_{\xi +\gamma }^{\left( \mu \right)
}\right) .
\end{array}
\label{recurrent-rel}
\end{equation}
The recursion is governed by the set $\Gamma _{\af\rightarrow \gf}$ called the injection fan. The latter is defined by the
carrier set $\left\{ \xi \right\} _{\af\rightarrow \gf}$ for the
coefficient function $s(\xi )$
\begin{equation*}
\left\{ \xi \right\} _{\widetilde{\af}\rightarrow \gf}:=\left\{
\xi \in P_{\widetilde{\af}}|s(\xi )\neq 0\right\}
\end{equation*}
appearing in the expansion
\begin{equation}
\prod_{\alpha \in \Delta ^{+}\setminus \Delta _{\bot }^{+}}\left( 1-e^{-\pi
_{\widetilde{\af}}\alpha }\right) ^{\mathrm{mult}(\alpha )-\mathrm{mult}%
_{\af}(\pi _{\widetilde{\af}}\alpha )}=-\sum_{\gamma \in P_{%
\widetilde{\af}}}s(\gamma )e^{-\gamma };\quad
\end{equation}
The weights in $\left\{ \xi \right\} _{\widetilde{\af}\rightarrow \gf}$ are to be shifted by $\gamma _{0}$ -- the lowest vector in $\left\{ \xi
\right\} $ -- and the zero element is to be eliminated:
\begin{equation}
\Gamma _{\af\rightarrow \gf}=\left\{ \xi -\gamma
_{0}|\xi \in \left\{ \xi \right\} \right\} \setminus \left\{ 0\right\} .
\end{equation}
Formula (\ref{eq:14}) is a particular case of recurrent relation for branching coefficients (\ref{recurrent-rel}) in the case of Cartan subalgebra $\af=\hfg$.

@@

Another particular case of this formula is connected with tensor
product decompositions. Consider the tensor product of two
irreducible $\gf$-modules $L^{\mu}\otimes L^{\nu}$. It is also a
$\gf$-module but not irreducible in general. So
\begin{equation}
  \label{eq:19}
  L^{\mu}\otimes L^{\nu}=\bigoplus_{\gamma} f^{\mu\nu}_{\gamma}L^{\gamma}
\end{equation}
The coefficients $f^{\mu\nu}_{\gamma}$ are called fusion
coefficients. The problem of computation of fusion coefficients is
equivalent to branching problem for the diagonal subalgebra
$\gf\subset \gf\oplus \gf$ (see \cite{LyakhovskyPostnova2011}). So our implementation of recurrent algorithm can be used to decompose tensor products (See Section \ref{sec:tens-prod-decomp}).

In the case of affine Lie algebras $\gf, \af$  the multiplicities
$m_{\nu}$ and the corresponding branching coefficients $b_{\nu}$
can be regarded as the coefficients in power series decomposition
of string and branching functions correspondingly:
\begin{align}
  \label{eq:21}
  &\sigma_{\nu}(q)=\sum_{n=0}^{\infty} m_{\nu-n\delta} q^n, \quad \nu=\sum_j c_j \omega_j,\quad c_j\geq 0\\
  & b_{\nu}(q)=\sum_{n=0}^{\infty} b_{\nu-n\delta} q^n,\quad  \nu=\sum_j c_j \omega_j, \quad c_j\geq 0
\end{align}
String and branching functions have  modular and analytic properties which are important for conformal field theory, especially in  coset models and the study of CFT on higher genus surfaces \cite{kac1988modular}, \cite{difrancesco1997cft}, \cite{Walton:1999xc}, \cite{walton1989conformal}.

{\bf Affine.m}  calculates weight multiplicities and branching coefficients for affine Lie algebras up to some finite grade. We present examples of computations in Sections \ref{sec:string-funct-affine}, \ref{sec:branch-funct-coset}. Now we proceed to the description of datastructures and algorithms implemented in {\bf Affine.m}.

\section{Core datastructures}
\label{sec:core-datastructures} Having introduced necessary
mathematical objects, problems and relations we now describe the
related datastructures of {\bf Affine.m}. Although {\it
Mathematica} is untyped language it is possible to create
structured objects and do type checks with patterns
\cite{shifrinmathematica}, \cite{maeder2000computer}.
\subsection{Weights}
\label{sec:weights}

Weights are represented  by two data-structures: \lstinline{finiteWeight} for finite-dimensional Lie algebras and \lstinline{affineWeight}  for affine.

@@

Affine weight is an extension of a finite weight by supplying it
with level and grade coordinates. There is a set of functions
defined for finite and affine weights. The complete list can be
found in  online help of the package. The most important
are definitions of addition, multiplication by a number and a
scalar product (bilinear form) for weights. These definitions
allow to use traditional notation when working with {\bf
Affine.m}:
\begin{lstlisting}
  w=makeFiniteWeight[{1,0,3}];
  v=makeFiniteWeight[{3,2,1}];
  2*w+v==makeFiniteWeight[{5,2,7}]
  w.v==6
\end{lstlisting}

The use of orthogonal basis in the internal structure of weights allows us to work with weights without complete specification of root system which is useful for study of branching, since roots of the subalgebra can be specified by hand.

@@

To specify an algebra of finite or affine type it is enough to fix
its root system. Root systems are represented by two datatypes
\lstinline{finiteRootSystem} and \lstinline{affineRootSystem}. The
latter is an extension of the former. We offer several different
constructors for these datastructures. It is possible to specify
the set of simple roots explicitly, for example to study the
subalgebra $B_2\subset B_4$ we can use the definition
\begin{lstlisting}
  b2b4=makeFiniteRootSystem[ { {1,-1,0,0}, {0,1,0,0} } ]
\end{lstlisting}
There are constructors for the root systems of simple finite-dimensional Lie algebras:
\begin{lstlisting}
  b2=makeSimpleRootSystem[B,2]
\end{lstlisting}
We use typographic features of {\it Mathematica} frontend to offer mathematical notation for simple Lie algebras:

@@

Non-twisted affine root systems can be created as affine extensions of finte root systems, e.g.
\begin{lstlisting}
  b2affine = makeAffineExtension[b2]
\end{lstlisting}
In notebook interface this can be written simply as $\hat{B}_2$.

Semisimple Lie algebras can be created as sums of simple:
\begin{lstlisting}[mathescape=true]
  $A_1\oplus A_1$ == finiteRootSystem[2, 2, {finiteWeight[2, {1, 0}], finiteWeight[2, {0, 1}]}]
\end{lstlisting}
%% The problem is here

Predicate \lstinline{rootSystemQ} checks if the object is root systems of finite or affine type.

List of simple roots is the property of root system so it is accessed as \lstinline{rs[simpleRoots]}.

We have implemented several functions to get major properties of root systems. Weyl vector is given by the function \lstinline{rho[rs_?rootSystemQ]}:
\begin{lstlisting}[label=list:1]
  In[1]  =  rho[b2]
  Out[1]  =  finiteWeight[2, {3/2, 1/2}]
\end{lstlisting}
Positive roots can be constructed with \lstinline{positiveRoots[rs_?rootSystemQ]}. For an affine Lie algebra this and related functions return the list up to fixed grade. This grade limit is set through the value of \lstinline{rs[gradeLimit]} which is equal to 10 by default. List of roots (up to \lstinline{gradeLimit}) is returned by \lstinline{roots[rs]}. Cartan matrix and fundamental weights are calculated by the functions \lstinline{cartanMatrix} and \lstinline{fundamentalWeights} correspondingly.

It is possible to specify a weight of a Lie algebra by its Dynkin labels
\begin{lstlisting}
  weight[b2][1,2] == makeFiniteWeight[{2, 1}]
\end{lstlisting}
The function \lstinline{dynkinLabels[rs_?rootSystemQ][wg_?weightQ]} returns Dynkin labels of weight \lstinline{wg} in the root system \lstinline{rs}.

Elements of Weyl group are specified by the set of indices of reflections, so the element  $w=s_{1}s_{2}s_{1}$ of the Weyl group of algebra $B_{2}$ is constructed with \lstinline{weylGroupElement[b2][1,2,1]}. Then it can be applied to the weights:
\begin{lstlisting}
  w = weylGroupElement[b2][1,2,1];
  w @ makeFiniteWeight[{1,0}] == makeFiniteWeight[{-1,0}]
\end{lstlisting}

Computation of lexicographically minimal form \cite{casselman1994machine,casselman1995automata} for Weyl group elements can be conveniently implemented through the use of pattern-matching in {\it Mathematica}. In \cite{KallenShortlex} rewrite rules for simple finite dimensional and affine Lie algebras are presented as {\it Mathematica} patterns. Our presentation of Weyl group elements is compatible with the code of \cite{KallenShortlex}
\begin{lstlisting}[mathescape=true]
  In[1]  = $<<$A3reduce;
           reduce[s[1,2,1,2,1,3,2,1,1]]
  Out[1] = s[2, 3, 2]

@@

We represent formal characters of representations by special structure \lstinline{formalElement}. This structure is a hash-table implemented with \lstinline{DownValues}. The keys are weights at the exponents and the values are the  corresponding multiplicities. \lstinline[mathescape=true]!makeFormalElement[{$\gamma_{1},\dots,\gamma_{n}$},{$m_{1},\dots,m_{n}$}]! creates data-structure which represents the element $\sum_{i=1}^{n} m_{i} e^{\gamma_{i}}$ of the formal algebra $\mathcal{E}$. The operations in $\mathcal{E}$ are implemented for \lstinline{formalElement} data-type: formal elements can be added, multiplied by a number or by an exponent of a weight. There exists also a multiplication of formal elements but no division.
\begin{lstlisting}[mathescape=true]
  In[1]  = makeFormalElement[{makeFiniteWeight[{1,1}],makeFiniteWeight[{0,0}]},{1,2}] *
             (2 * Exp[makeFiniteWeight[{1,0}]] *
             makeFormalElement[{makeFiniteWeight[{1,1}],makeFiniteWeight[{0,0}]},{1,2}]);
  In[2]  = In[1][weights]
  Out[2] = {finiteWeight[2, {1, 0}], finiteWeight[2, {2, 1}], finiteWeight[2, {3, 2}]}

@@

{\bf Affine.m} can be used to study different kinds of modules, i.e. Verma modules, irreducible modules and parabolic Verma modules.  We need datastructure \lstinline{module} to represent generic module of Lie algebra $\gf$. Module properties can be deduced from its set of singular weights using Weyl character formulae \eqref{eq:11},\eqref{eq:12},\eqref{eq:18},\eqref{eq:13}. Set of singular weights can have Weyl symmetry. It can be a symmetry with respect to the Weyl group $W_{\gf}$ or with respect to some subalgebra $W_{\af}$ as in the case of parabolic Verma modules. Then it is possible to study only the main Weyl chamber $C_{\af}$. To use this symmetry generic constructor for \lstinline{module} datastructure accepts several parameters \lstinline{makeModule[rs_?rootSystemQ][singWeights_formalElement,subs_?rootSystemQ|emptyRootSystem[],limit:10}. Here \lstinline{rs} is the root system of the Lie algebra $\gf$,\lstinline{singWeights} is the set of singular weights,  \lstinline{subs} is the root system corresponding to the Weyl group $W_{\af}$ which is the (anti-)symmetry of the set of singular weights. Parameter \lstinline{limit} limits computation for infinite-dimensional modules such as Verma or parabolic Verma.
There are several specialized constructors for different types of highest weight modules:
\begin{lstlisting}[mathescape=true]
vm=makeVermaModule[$B_{2}$][{2,1}];
pm=makeParabolicVermaModule[$B_{2}$][weight[$B_{2}$][2,1],{1}];
im=makeIrreducibleModule[$B_{2}$][2,1];
GraphicsRow[textPlot/@{im,vm,pm}]
$\includegraphics[width=130mm]{figures/irrep-verma-pverma.pdf}$
\end{lstlisting}

As we already stated properties of a module are encoded by its singular element. Function \lstinline{singularElement[m_module]} returns singular element of a module as \lstinline{formalElement} datastructure. Character (up to \lstinline{limit} for (parabolic) Verma modules) is returned by the function \lstinline{character[m_module]}. Direct sum of modules is a module and we use natural notation
\begin{lstlisting}[mathescape=true]
im1=makeIrreducibleModule[$B_{2}$][weight[$B_{2}$][2,1]];
im2=makeIrreducibleModule[$B_{2}$][weight[$B_{2}$][1,2]];
textPlot[im1$\oplus$ im2]
$\includegraphics[width=60mm]{figures/irrep-sum.pdf}$
\end{lstlisting}

The tensor product is also implemented but only for the finite-dimensional Lie algebras, since tensor product of affine Lie algebra modules leads to rich new structures \cite{kazhdan1994tensor3,kazhdan1993tensor1,kazhdan1993tensor2} which are out of the scope of the present paper.
\begin{lstlisting}[mathescape=true]
textPlot[makeIrreducibleModule[$A_{1}$][5]$\otimes$ makeIrreducibleModule[$A_{1}$][3]];
$\includegraphics[width=60mm]{figures/tensor-product.pdf}$
\end{lstlisting}
\section{Computational algorithms}
\label{sec:comp-algor}

As we have already stated in section \ref{sec:high-weight-modul} there exist two recurrent relations which can be used to calculate weight multiplicities in irreducible modules. Both algorithms proceed in the following way to calculate weight multiplicities:
\begin{enumerate}
\item Create the list of weights in the main Weyl chamber by subtracting all possible combinations of simple roots from the highest weight (e.g. for a finite-dimensional algebra subtract $\alpha_{1}$ from $\mu$ while inside $\bar C$, then subtract $\alpha_{2}$ from all the weights already obtained etc).
\item Sort the list of weights by their product with the Weyl vector.
\item Use a recurrent formula. If the weight required for recurrent computation is outside the main chamber use the Weyl symmetry.
\end{enumerate}
The difference in the performance of algorithms is in the number of previous values required to get the multiplicity of weight under consideration. For  the Weyl formula based recurrent relation \eqref{eq:14} it is constant and equal to the number of elements in the Weyl group (if we are far from the boundary of representation diagram). When the Freudenthal formula \eqref{eq:15} is used number of previous values grows with the distance from the external border of representation. So the Freudenthal formula is faster if the weight is close to the border or the rank of the algebra and the size of Weyl group is large \cite{moody1982fast}.
Note that the Freudenthal formula is valid for irreducible modules only, so it can not be used to study (generalized) Verma modules.

We have made some experiments with our implementations of the Freudenthal formula and formula \eqref{eq:15} and have got Figure  \ref{fig:freudenthal-racah-times}, which depicts the dependence of the computation time on the number of weights in a module.

>>>>

Consider an action of the element $s_{\alpha}s_{\alpha+\delta}$ of the Weyl group of affine Lie algebra  $\hat\gf$ for $\alpha$ being a simple root of the underlying finite-dimensional Lie algebra $\gf$. Using definition \eqref{eq:8} it is easy to see that $s_{\alpha}s_{\alpha+\delta} \cdot \lambda=\lambda+\frac{2}{(\alpha,\alpha)}\alpha+\left(\frac{(\alpha,\alpha)}{2 (\lambda,\delta)}+(\lambda,\alpha)\right) \delta$. So the Weyl group $W_{\hat\gf}$ can be presented as a semidirect product of the Weyl group $W_{\gf}$ of $\gf$ and the set of translations corresponding to the roots of $\gf$.

@@

For a subalgebra $\af\subset \gf$ we introduce the subalgebra
$\afb$. Consider a root subspace $\hf_{\perp \af}^{\ast }$
orthogonal to $\hf_{\af}$,
\begin{equation*}
\hf_{\perp \af}^{\ast }:=\left\{ \eta \in \hf^{\ast }|\forall
h\in \hf_{\af};\eta \left( h\right) =0\right\} ,
\end{equation*}
and the roots (correspondingly -- positive roots) of $\gf$ orthogonal
to the roots of $\af$,
\begin{eqnarray}
\Delta _{\afb} &:&=\left\{ \beta \in \Delta _{\gf}|\forall
h\in \hf_{\af};\beta \left( h\right) =0\right\} ,
\label{delta a ort} \\
\Delta _{\afb}^{+} &:&=\left\{ \beta ^{+}\in \Delta _{\gf%
}^{+}|\forall h\in \hf_{\af};\beta ^{+}\left( h\right) =0\right\} .
\notag
\end{eqnarray}
Let $W_{\afb}$ be a subgroup of $W$ generated by
the reflections $w_{\beta }$ with the roots $\beta \in \Delta _{\af_{\perp
}}^{+}$. The subsystem $\Delta _{\afb}$ determines a 
subalgebra $\afb$ with the Cartan subalgebra $\hf_{\af%
_{\perp }}$.

@@

Denote by  $k_{\xi }^{\left( \mu \right) }$ signed branching coefficients defined as follows. If $\xi\in \bar C_{\af}$ is in the main Weyl chamber $k_{\xi}^{(\mu)}=b^{(\mu)}_{\xi}$, otherwise $k_{\xi}^{(\mu)}=\epsilon(w) b^{(\mu)}_{w (\xi+\rho_{\af})-\rho_{\af}}$ where $w\in W_{\af}$ is such that $w (\xi+\rho_{\af})-\rho_{\af}\in \bar C_{\af}$.

Now we can use the Weyl character formula to write a
recurrent relation \cite{2010arXiv1007.0318L} for the signed branching
coefficients $k_{\xi }^{\left( \mu \right) }$ corresponding to an
injection $\af\hookrightarrow \gf$:
\begin{equation}
\begin{array}{c}
k_{\xi }^{\left( \mu \right) }=-\frac{1}{s\left( \gamma _{0}\right) }\left(
\sum_{u\in U}\epsilon (u)\;\dim \left( L_{\afb}^{\mu _{\af%
_{\perp }}\left( u\right) }\right) \delta _{\xi -\gamma _{0},\pi _{%
\widetilde{\af}}(u(\mu +\rho )-\rho )}+\right.  \\
\left. +\sum_{\gamma \in \Gamma _{\widetilde{\af}\rightarrow \gf%
}}s\left( \gamma +\gamma _{0}\right) k_{\xi +\gamma }^{\left( \mu \right)
}\right) .
\end{array}
\label{recurrent-rel}
\end{equation}
The recursion is governed by the set $\Gamma _{\af\rightarrow \gf}$ called the injection fan. The latter is defined by the
carrier set $\left\{ \xi \right\} _{\af\rightarrow \gf}$ for the
coefficient function $s(\xi )$
\begin{equation*}
\left\{ \xi \right\} _{\widetilde{\af}\rightarrow \gf}:=\left\{
\xi \in P_{\widetilde{\af}}|s(\xi )\neq 0\right\}
\end{equation*}
appearing in the expansion
\begin{equation}
\prod_{\alpha \in \Delta ^{+}\setminus \Delta _{\bot }^{+}}\left( 1-e^{-\pi
_{\widetilde{\af}}\alpha }\right) ^{\mathrm{mult}(\alpha )-\mathrm{mult}%
_{\af}(\pi _{\widetilde{\af}}\alpha )}=-\sum_{\gamma \in P_{%
\widetilde{\af}}}s(\gamma )e^{-\gamma };\quad
\end{equation}
The weights in $\left\{ \xi \right\} _{\widetilde{\af}\rightarrow \gf}$ are to be shifted by $\gamma _{0}$ -- the lowest vector in $\left\{ \xi
\right\} $ -- and the zero element is to be eliminated:
\begin{equation}
\Gamma _{\af\rightarrow \gf}=\left\{ \xi -\gamma
_{0}|\xi \in \left\{ \xi \right\} \right\} \setminus \left\{ 0\right\} .
\end{equation}
The formula (\ref{eq:14}) is a particular case of the recurrent relation for branching coefficients (\ref{recurrent-rel}) in the case of a Cartan subalgebra $\af=\hfg$.

@@

Another particular case of this formula is connected with tensor
product decompositions. Consider the tensor product of two
irreducible $\gf$-modules $L^{\mu}\otimes L^{\nu}$. It is also a
$\gf$-module but not irreducible in general. So
\begin{equation}
  \label{eq:19}
  L^{\mu}\otimes L^{\nu}=\bigoplus_{\gamma} f^{\mu\nu}_{\gamma}L^{\gamma}
\end{equation}
The coefficients $f^{\mu\nu}_{\gamma}$ are called the fusion
coefficients. The problem of computation of fusion coefficients is
equivalent to a branching problem for the diagonal subalgebra
$\gf\subset \gf\oplus \gf$ (see \cite{LyakhovskyPostnova2011}). So our implementation of a recurrent algorithm can be used to decompose tensor products (See Section \ref{sec:tens-prod-decomp}).

In a case of affine Lie algebras $\gf, \af$  the multiplicities
$m_{\nu}$ and the branching coefficients $b^{(\mu)}_{\nu}$
can be regarded as the coefficients in the power series decomposition
of string and branching functions correspondingly:
\begin{align}
  \label{eq:21}
  &\sigma_{\nu}(q)=\sum_{n=0}^{\infty} m_{\nu-n\delta} q^n, \quad \nu=\sum_j c_j \omega_j,\quad c_j\geq 0\\
  & b_{\nu}(q)=\sum_{n=0}^{\infty} b_{\nu-n\delta} q^n,\quad  \nu=\sum_j c_j \omega_j, \quad c_j\geq 0
\end{align}
String and branching functions have  modular and analytic properties which are important for conformal field theory, especially in  coset models and the study of CFT on higher genus surfaces \cite{kac1988modular}, \cite{difrancesco1997cft}, \cite{Walton:1999xc}, \cite{walton1989conformal}.

{\bf Affine.m}  calculates weight multiplicities and branching coefficients for affine Lie algebras up to some finite grade. We present examples of computations in Sections \ref{sec:string-funct-affine}, \ref{sec:branch-funct-coset}. Now we proceed to the description of the datastructures and the algorithms implemented in {\bf Affine.m}.

\section{Core datastructures}
\label{sec:core-datastructures} Having introduced necessary
mathematical objects, problems and relations we now describe the
related datastructures of {\bf Affine.m}. Although {\it
Mathematica} is untyped language it is possible to create
structured objects and do the type checks with patterns
\cite{shifrinmathematica}, \cite{maeder2000computer}.
\subsection{Weights}
\label{sec:weights}

Weights are represented  by two datastructures: \lstinline{finiteWeight} for finite-dimensional Lie algebras and \lstinline{affineWeight}  for affine.

@@

An affine weight is an extension of a finite weight by supplying it
with the level and grade coordinates. There is a set of functions
defined for finite and affine weights. The complete list can be
found in the  online help of the package. The most important
are the definitions of an addition, a multiplication by a number and a
scalar product (bilinear form) for weights. These definitions
allow us to use a traditional notation with {\bf Affine.m}:
\begin{lstlisting}
  w=makeFiniteWeight[{1,0,3}];
  v=makeFiniteWeight[{3,2,1}];
  2*w+v==makeFiniteWeight[{5,2,7}]
  w.v==6
\end{lstlisting}

The use of an orthogonal basis in the internal structure of weights allows us to work with weights without a complete specification of a root system which is useful for a study of branching, since the roots of the subalgebra can be specified by hand.

@@

To specify an algebra of finite or affine type it is enough to fix
its root system. Root systems are represented by two datatypes
\lstinline{finiteRootSystem} and \lstinline{affineRootSystem}. The
latter is an extension of the former. We offer several different
constructors for these datastructures. It is possible to specify
the set of simple roots explicitly, for example to study the
subalgebra $B_2\subset B_4$ we can use the definition
\begin{lstlisting}
  b2b4=makeFiniteRootSystem[ { {1,-1,0,0}, {0,1,0,0} } ]
\end{lstlisting}
There are constructors for the root systems of simple finite-dimensional Lie algebras:
\begin{lstlisting}
  b2=makeSimpleRootSystem[B,2]
\end{lstlisting}
We use the typographic features of the {\it Mathematica} frontend to offer a mathematical notation for simple Lie algebras:

@@

Non-twisted affine root systems can be created as the affine extensions of finte root systems, e.g.
\begin{lstlisting}
  b2affine = makeAffineExtension[b2]
\end{lstlisting}
In the notebook interface this can be written simply as $\hat{B}_2$.

Semisimple Lie algebras can be created as the sums of simple ones:
\begin{lstlisting}[mathescape=true]
  $A_1\oplus A_1$ == finiteRootSystem[2, 2, {finiteWeight[2, {1, 0}], finiteWeight[2, {0, 1}]}]
\end{lstlisting}
%% The problem is here

The predicate \lstinline{rootSystemQ} checks if the object is a root systems of finite or affine type.

The list of simple roots is the property of root system so it is accessed as the field \lstinline{rs[simpleRoots]}.

We have implemented several functions to get major properties of root systems. Weyl vector is computed by the function \lstinline{rho[rs_?rootSystemQ]}:
\begin{lstlisting}[label=list:1]
  In[1]  =  rho[b2]
  Out[1]  =  finiteWeight[2, {3/2, 1/2}]
\end{lstlisting}
The list of positive roots can be constructed with the function \lstinline{positiveRoots[rs_?rootSystemQ]}. For an affine Lie algebra this and related functions return the list up to some fixed grade. This grade limit is set as the value of the field \lstinline{rs[gradeLimit]} which is equal to 10 by default. The list of roots (up to \lstinline{gradeLimit}) is returned by the function \lstinline{roots[rs]}. Cartan matrix and the fundamental weights are calculated by the functions \lstinline{cartanMatrix} and \lstinline{fundamentalWeights} correspondingly.

It is possible to specify a weight of a Lie algebra by its Dynkin labels
\begin{lstlisting}
  weight[b2][1,2] == makeFiniteWeight[{2, 1}]
\end{lstlisting}
The function \lstinline{dynkinLabels[rs_?rootSystemQ][wg_?weightQ]} returns Dynkin labels of a weight \lstinline{wg} in the root system \lstinline{rs}.

An element of the Weyl group can be specified by the set of the indices of the reflections, so the element  $w=s_{1}s_{2}s_{1}$ of the Weyl group of algebra $B_{2}$ is constructed with \lstinline{weylGroupElement[b2][1,2,1]}. Then it can be applied to the weights:
\begin{lstlisting}
  w = weylGroupElement[b2][1,2,1];
  w @ makeFiniteWeight[{1,0}] == makeFiniteWeight[{-1,0}]
\end{lstlisting}

A computation of a lexicographically minimal form \cite{casselman1994machine,casselman1995automata} for the Weyl group elements can be conveniently implemented using pattern-matching in {\it Mathematica}. In \cite{KallenShortlex} the rewrite rules for simple finite dimensional and affine Lie algebras are presented as the {\it Mathematica} patterns. Our presentation of the Weyl group elements is compatible with the code of \cite{KallenShortlex}:
\begin{lstlisting}[mathescape=true]
  In[1]  = $<<$A3reduce;
           reduce[s[1,2,1,2,1,3,2,1,1]]
  Out[1] = s[2, 3, 2]

@@

We represent the formal characters of modules by a special structure \lstinline{formalElement}.  The constructor \lstinline[mathescape=true]!makeFormalElement[{$\gamma_{1},\dots,\gamma_{n}$},{$m_{1},\dots,m_{n}$}]! creates a datastructure which represents the element $\sum_{i=1}^{n} m_{i} e^{\gamma_{i}}$ of the formal algebra $\mathcal{E}$. This structure is a hash-table implemented with \lstinline{DownValues}. The keys are weights at the exponents and the values are the  corresponding multiplicities. The operations in $\mathcal{E}$ are implemented for the \lstinline{formalElement} data-type: formal elements can be added, multiplied by a number or by an exponent of a weight. There exists also a multiplication of formal elements but no division.
\begin{lstlisting}[mathescape=true]
  In[1]  = makeFormalElement[{makeFiniteWeight[{1,1}],makeFiniteWeight[{0,0}]},{1,2}] *
             (2 * Exp[makeFiniteWeight[{1,0}]] *
             makeFormalElement[{makeFiniteWeight[{1,1}],makeFiniteWeight[{0,0}]},{1,2}]);
  In[2]  = In[1][weights]
  Out[2] = {finiteWeight[2, {1, 0}], finiteWeight[2, {2, 1}], finiteWeight[2, {3, 2}]}

@@

{\bf Affine.m} can be used to study different kinds of modules, i.e. Verma modules, irreducible modules and parabolic Verma modules.  We need the datastructure \lstinline{module} to represent a generic module of a Lie algebra $\gf$. Module properties can be deduced from its set of singular weights using the Weyl character formulae \eqref{eq:11},\eqref{eq:12},\eqref{eq:18},\eqref{eq:13}. A set of singular weights can have Weyl symmetry. It can be a symmetry with respect to the Weyl group $W_{\gf}$ or with respect to some subgroup $W_{\af}$ as in the case of parabolic Verma modules. Then it is possible to study only the main Weyl chamber $C_{\af}$. To use this symmetry a generic constructor for the \lstinline{module} datastructure accepts several parameters \lstinline{makeModule[rs_?rootSystemQ][singWeights_formalElement,subs_?rootSystemQ|emptyRootSystem[],limit:10}. Here \lstinline{rs} is the root system of a Lie algebra $\gf$, \;\lstinline{singWeights} is the set of singular weights,  \lstinline{subs} is the root system corresponding to the Weyl group $W_{\af}$ which is the (anti-)symmetry of the set of singular weights. The parameter \lstinline{limit} limits the computation for infinite-dimensional representations such as Verma or parabolic Verma modules.
There are several specialized constructors for different types of highest weight modules:
\begin{lstlisting}[mathescape=true]
vm=makeVermaModule[$B_{2}$][{2,1}];
pm=makeParabolicVermaModule[$B_{2}$][weight[$B_{2}$][2,1],{1}];
im=makeIrreducibleModule[$B_{2}$][2,1];
GraphicsRow[textPlot/@{im,vm,pm}]
$\includegraphics[width=130mm]{figures/irrep-verma-pverma.pdf}$
\end{lstlisting}

As we have already stated the properties of a module are encoded by its singular element. The function \lstinline{singularElement[m_module]} returns the singular element of a module as a \lstinline{formalElement} datastructure. The character (up to \lstinline{limit} for (parabolic) Verma modules) is returned by the function \lstinline{character[m_module]}. A direct sum of modules is a module and we use natural notation
\begin{lstlisting}[mathescape=true]
In[1] := im1=makeIrreducibleModule[$B_{2}$][weight[$B_{2}$][2,1]];
         im2=makeIrreducibleModule[$B_{2}$][weight[$B_{2}$][1,2]];
         Head[im1$\oplus$ im2]
Out[1] = module
In[2] := textPlot[im1$\oplus$ im2]
$\includegraphics[width=60mm]{figures/irrep-sum.pdf}$
\end{lstlisting}

The tensor product is also implemented but only for finite-dimensional Lie algebras, since tensor product of affine Lie algebra modules leads to rich new structures \cite{kazhdan1994tensor3,kazhdan1993tensor1,kazhdan1993tensor2} which are out of the scope of the present paper.
\begin{lstlisting}[mathescape=true]
textPlot[makeIrreducibleModule[$A_{1}$][5]$\otimes$ makeIrreducibleModule[$A_{1}$][3]];
$\includegraphics[width=60mm]{figures/tensor-product.pdf}$
\end{lstlisting}
\section{Computational algorithms}
\label{sec:comp-algor}

As we have already stated in the section \ref{sec:high-weight-modul} there exist two recurrent relations which can be used to calculate weight multiplicities in irreducible modules. Both algorithms proceed in the following way to calculate weight multiplicities:
\begin{enumerate}
\item Create the list of weights in the main Weyl chamber by subtracting all possible combinations of simple roots from the highest weight (e.g. for a finite-dimensional algebra subtract $\alpha_{1}$ from $\mu$ while inside $\bar C$, then subtract $\alpha_{2}$ from all the weights already obtained etc).
\item Sort the list of weights by their product with the Weyl vector.
\item Use a recurrent formula. If the weight required for the recurrent computation is outside the main chamber use the Weyl symmetry.
\end{enumerate}
The difference in the performance of algorithms comes from the number of previous values required to compute the multiplicity of a weight under consideration. For  the  recurrent relation \eqref{eq:14} based on the Weyl formula it is constant and equal to the number of elements in the Weyl group (if we are far from the boundary of representation diagram). When the Freudenthal formula \eqref{eq:15} is used the number of the previous values grows with the distance from the external border of representation. So the Freudenthal formula is faster if the weight is close to the border or the rank of the algebra and the size of Weyl group is large \cite{moody1982fast}.
Note that the Freudenthal formula is valid for the irreducible modules only, so it can not be used to study (generalized) Verma modules.

We have made some experiments with our implementations of the Freudenthal formula and formula \eqref{eq:15} and prepared the Figure  \ref{fig:freudenthal-racah-times}, which depicts the dependence of the computation time on the number of weights in a module.

\begin{figure}[h]
  \noindent\centering{
    \includegraphics[width=80mm]{figures/timing.pdf}
  }
  \caption{The dependence of the running time of the algorithms based on the Freudenthal formula \eqref{eq:15} (dashed) and the recurrent relation \eqref{eq:14} (solid) on the number of weights in $\bar C$ for calculation of multiplicities in representations of $B_{2}$.}
\label{fig:freudenthal-racah-times}
